---
title: "Gov 50 Final Project"
author: "Chase Yager"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

I am interested in exploring data related to rates of COVID-19 infection, detection, treatment, and death on the basis of 
socioeconomic class in the United States. The focus will be on whether being located in a different socioeconomic class 
has a negative impact on the rates of detection and treatment of an illness such as COVID-19, and further on whether or 
not the rates of death when an illness is detected is altered based on this same metric. The ultimate goal would be to 
determine if there is an effect, and then explore possible explanations for that effect.

The question which I am seeking to answer with my research is what the impact of poverty levels are on the incidence and 
mortality rates of COVID-19. My hypothesis is that areas with higher poverty will experience higher case counts and mortality 
rates due to a lack of access to the high quality healthcare and lower rates of insurance. The explanatory variable will be 
poverty rates by county in the US. The outcome variable will be incidence of COVID-19 and number of deaths. If the incidence 
rate and death rate are higher in counties with higher poverty, then this would support our hypothesis. If the rates are the 
same, however, this would disprove our hypothesis and indicate equivalent access to care among poor and wealthy Americans.

## Data Assembly
```{r}
library(tidyverse)
library(ggplot2)
library(broom)
```

```{r}
cases <- read.csv("data/covid_confirmed_usafacts.csv")
deaths <- read.csv("data/covid_deaths_usafacts.csv")
poverty <- read.csv("data/poverty_by_county.csv")
```
```{r}
fin_cases <- cases |>
  select(countyFIPS, County.Name, X2022.10.14) |>
  mutate(final_count = X2022.10.14) |>
  select(countyFIPS, County.Name, final_count)
```
```{r}
fin_deaths <- deaths |>
  select(countyFIPS, County.Name, X2022.10.14) |>
  mutate(final_deaths = X2022.10.14) |>
  select(countyFIPS, County.Name, final_deaths)
```

```{r}
count_deaths <- fin_cases |>
  inner_join(fin_deaths, by = "countyFIPS") |>
  mutate(County_Name = County.Name.x) |>
  select(countyFIPS, County_Name, final_count, final_deaths)
```

```{r}
poverty <- poverty |>
  mutate(X = as.numeric(X))
```
```{r}
count_deaths <- count_deaths |>
  mutate(countyFIPS = as.numeric(countyFIPS))
```


```{r}
poverty$X[poverty$X.1 == "AL"] <- (poverty$X[poverty$X.1 == "AL"] + 1000) 
poverty$X[poverty$X.1 == "AK"] <- (poverty$X[poverty$X.1 == "AK"] + 2000)
poverty$X[poverty$X.1 == "AZ"] <- (poverty$X[poverty$X.1 == "AZ"] + 4000) 
poverty$X[poverty$X.1 == "AR"] <- (poverty$X[poverty$X.1 == "AR"] + 5000) 
poverty$X[poverty$X.1 == "CA"] <- (poverty$X[poverty$X.1 == "CA"] + 6000) 
poverty$X[poverty$X.1 == "CO"] <- (poverty$X[poverty$X.1 == "CO"] + 8000) 
poverty$X[poverty$X.1 == "CT"] <- (poverty$X[poverty$X.1 == "CT"] + 9000) 
poverty$X[poverty$X.1 == "DE"] <- (poverty$X[poverty$X.1 == "DE"] + 10000) 
poverty$X[poverty$X.1 == "DC"] <- (poverty$X[poverty$X.1 == "DC"] + 11000) 
poverty$X[poverty$X.1 == "FL"] <- (poverty$X[poverty$X.1 == "FL"] + 12000) 
poverty$X[poverty$X.1 == "GA"] <- (poverty$X[poverty$X.1 == "GA"] + 13000) 
poverty$X[poverty$X.1 == "HI"] <- (poverty$X[poverty$X.1 == "HI"] + 15000) 
poverty$X[poverty$X.1 == "ID"] <- (poverty$X[poverty$X.1 == "ID"] + 16000) 
poverty$X[poverty$X.1 == "IL"] <- (poverty$X[poverty$X.1 == "IL"] + 17000) 
poverty$X[poverty$X.1 == "IN"] <- (poverty$X[poverty$X.1 == "IN"] + 18000) 
poverty$X[poverty$X.1 == "IA"] <- (poverty$X[poverty$X.1 == "IA"] + 19000) 
poverty$X[poverty$X.1 == "KS"] <- (poverty$X[poverty$X.1 == "KS"] + 20000) 
poverty$X[poverty$X.1 == "KY"] <- (poverty$X[poverty$X.1 == "KY"] + 21000) 
poverty$X[poverty$X.1 == "LA"] <- (poverty$X[poverty$X.1 == "LA"] + 22000) 
poverty$X[poverty$X.1 == "ME"] <- (poverty$X[poverty$X.1 == "ME"] + 23000) 
poverty$X[poverty$X.1 == "MD"] <- (poverty$X[poverty$X.1 == "MD"] + 24000) 
poverty$X[poverty$X.1 == "MA"] <- (poverty$X[poverty$X.1 == "MA"] + 25000)
poverty$X[poverty$X.1 == "MI"] <- (poverty$X[poverty$X.1 == "MI"] + 26000) 
poverty$X[poverty$X.1 == "MN"] <- (poverty$X[poverty$X.1 == "MN"] + 27000) 
poverty$X[poverty$X.1 == "MS"] <- (poverty$X[poverty$X.1 == "MS"] + 28000) 
poverty$X[poverty$X.1 == "MO"] <- (poverty$X[poverty$X.1 == "MO"] + 29000) 
poverty$X[poverty$X.1 == "MT"] <- (poverty$X[poverty$X.1 == "MT"] + 30000) 
poverty$X[poverty$X.1 == "NE"] <- (poverty$X[poverty$X.1 == "NE"] + 31000) 
poverty$X[poverty$X.1 == "NV"] <- (poverty$X[poverty$X.1 == "NV"] + 32000) 
poverty$X[poverty$X.1 == "NH"] <- (poverty$X[poverty$X.1 == "NH"] + 33000) 
poverty$X[poverty$X.1 == "NJ"] <- (poverty$X[poverty$X.1 == "NJ"] + 34000) 
poverty$X[poverty$X.1 == "NY"] <- (poverty$X[poverty$X.1 == "NY"] + 36000)
poverty$X[poverty$X.1 == "NC"] <- (poverty$X[poverty$X.1 == "NC"] + 37000) 
poverty$X[poverty$X.1 == "ND"] <- (poverty$X[poverty$X.1 == "ND"] + 38000)
poverty$X[poverty$X.1 == "OH"] <- (poverty$X[poverty$X.1 == "OH"] + 39000)
poverty$X[poverty$X.1 == "OK"] <- (poverty$X[poverty$X.1 == "OK"] + 40000)
poverty$X[poverty$X.1 == "OR"] <- (poverty$X[poverty$X.1 == "OR"] + 41000) 
poverty$X[poverty$X.1 == "PA"] <- (poverty$X[poverty$X.1 == "PA"] + 42000) 
poverty$X[poverty$X.1 == "RI"] <- (poverty$X[poverty$X.1 == "RI"] + 44000) 
poverty$X[poverty$X.1 == "SC"] <- (poverty$X[poverty$X.1 == "SC"] + 45000) 
poverty$X[poverty$X.1 == "SD"] <- (poverty$X[poverty$X.1 == "SD"] + 46000) 
poverty$X[poverty$X.1 == "TN"] <- (poverty$X[poverty$X.1 == "TN"] + 47000) 
poverty$X[poverty$X.1 == "TX"] <- (poverty$X[poverty$X.1 == "TX"] + 48000) 
poverty$X[poverty$X.1 == "UT"] <- (poverty$X[poverty$X.1 == "UT"] + 49000) 
poverty$X[poverty$X.1 == "VT"] <- (poverty$X[poverty$X.1 == "VT"] + 50000) 
poverty$X[poverty$X.1 == "VA"] <- (poverty$X[poverty$X.1 == "VA"] + 51000) 
poverty$X[poverty$X.1 == "WA"] <- (poverty$X[poverty$X.1 == "WA"] + 53000) 
poverty$X[poverty$X.1 == "WV"] <- (poverty$X[poverty$X.1 == "WV"] + 54000) 
poverty$X[poverty$X.1 == "WI"] <- (poverty$X[poverty$X.1 == "WI"] + 55000) 
poverty$X[poverty$X.1 == "WY"] <- (poverty$X[poverty$X.1 == "WY"] + 56000) 
```


```{r}
poverty <- poverty |>
  mutate(countyFIPS = X)
```

```{r}
poverty <- poverty |>
  subset(countyFIPS!=0)
```
```{r}
count_deaths <- count_deaths |>
  subset(countyFIPS != 0)
```

```{r}
dataset <- poverty |>
  inner_join(count_deaths)
```

```{r}
dataset <- dataset |>
  mutate(Poverty_Count = X.3, Poverty_Perc = as.numeric(X.6)) |>
  select(countyFIPS, County_Name, final_count, final_deaths, Poverty_Count, Poverty_Perc)
```
## Data Visualization
```{r}
pov_count_graph <- dataset |>
  ggplot(mapping = aes (x = Poverty_Perc, y = final_count)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue") +
  scale_y_log10() +
  labs(x = "Percent in Poverty by County", y = "Number of Confirmed Cases", title = "Comparison Between Poverty Rate and COVID Cases by County")
pov_count_graph
```
```{r}
pov_death_graph <- dataset |>
  ggplot(mapping = aes (x = Poverty_Perc, y = final_deaths)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue") +
  scale_y_log10() +
  labs(x = "Percent in Poverty by County", y = "Number of Confirmed Deaths", title = "Comparison Between Poverty Rate and COVID Deaths by County")

pov_death_graph
```

## Analysis

```{r}
dataset <- dataset |>
  mutate(per_fatal = 100*(final_deaths/final_count))
```
```{r}
perc_comp_fig <- dataset |>
  ggplot(mapping = aes(x = Poverty_Perc, y = per_fatal)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Percentage in Poverty Compared to Percentage of Cases that were Fatal", x = "Poverty Percentage", y = "Percentage of Cases that were Fatal")
perc_comp_fig
```
```{r}
fit <- lm(per_fatal ~ Poverty_Perc, data = dataset) 

fit |>
  broom::tidy() |>
  select(term, estimate) |>
  knitr::kable()
```
From the linear regression run with percentage of the population in poverty as the predictor and the percentage of 
cases that were fatal as the outcome variable, the results indicate an intercept of .719 and a slope of .054. This 
indicates that, in the case when there was zero poverty we would expect to see a percentage of fatal cases of .719%. 
Additionally, the slope indicates that for every 1% increase in the poverty percentage, we would see a .05% increase 
in the fatality rate. This is a relatively small relationship, however the positive relationship still supports the 
original hypothesis that places with a higher poverty rate would experience higher rates of covid fatalities. 

```{r}
fit.sum <- summary(fit)

```

