---
title: "Gov 50 Final Project"
author: "Chase Yager"
description: "COVID-19 Mortality Rates as a Result of the Limitations of Poverty"
output:
  distill::distill_article:
    self_contained: false
---



## Introduction

This study seeks to explore data pertaining to COVID-19 infection and death on the basis of socioeconomic class in
the United States. The question which this study seeks to answer through this research is what the impact of poverty
levels are on the incidence and mortality rates of COVID-19. 

My hypothesis is that areas with higher poverty will experience higher case counts and mortality rates due to a lack
of access to high-quality healthcare, decreased ability to take time off work due to the need for money, decreased
ability to work remotely and avoid exposure to the illness, and lower rates of insurance. The explanatory variable
will be poverty rates by county in the US. The outcome variable will mortality rates due to COVID, which will be
found by dividing deaths due to COVID by incidences of COVID in each county. If the incidence rate and death rate
are higher in counties with higher poverty, then this would support our hypothesis. If the rates are the same,
however, this would disprove our hypothesis and indicate equivalent access to care among poor and wealthy Americans.


## Data Assembly:
```{r}
library(tidyverse)
library(ggplot2)
library(broom)
```

```{r}
cases <- read.csv("data/covid_confirmed_usafacts.csv")
deaths <- read.csv("data/covid_deaths_usafacts.csv")
poverty <- read.csv("data/poverty_by_county.csv")
```
```{r}
fin_cases <- cases |>
  select(countyFIPS, County.Name, X2022.10.14) |>
  mutate(final_count = X2022.10.14) |>
  select(countyFIPS, County.Name, final_count)
```
```{r}
fin_deaths <- deaths |>
  select(countyFIPS, County.Name, X2022.10.14) |>
  mutate(final_deaths = X2022.10.14) |>
  select(countyFIPS, County.Name, final_deaths)
```

```{r}
count_deaths <- fin_cases |>
  inner_join(fin_deaths, by = "countyFIPS") |>
  mutate(County_Name = County.Name.x) |>
  select(countyFIPS, County_Name, final_count, final_deaths)
```

```{r}
poverty <- poverty |>
  mutate(X = as.numeric(X))
```
```{r}
count_deaths <- count_deaths |>
  mutate(countyFIPS = as.numeric(countyFIPS))
```


```{r}
poverty$X[poverty$X.1 == "AL"] <- (poverty$X[poverty$X.1 == "AL"] + 1000) 
poverty$X[poverty$X.1 == "AK"] <- (poverty$X[poverty$X.1 == "AK"] + 2000)
poverty$X[poverty$X.1 == "AZ"] <- (poverty$X[poverty$X.1 == "AZ"] + 4000) 
poverty$X[poverty$X.1 == "AR"] <- (poverty$X[poverty$X.1 == "AR"] + 5000) 
poverty$X[poverty$X.1 == "CA"] <- (poverty$X[poverty$X.1 == "CA"] + 6000) 
poverty$X[poverty$X.1 == "CO"] <- (poverty$X[poverty$X.1 == "CO"] + 8000) 
poverty$X[poverty$X.1 == "CT"] <- (poverty$X[poverty$X.1 == "CT"] + 9000) 
poverty$X[poverty$X.1 == "DE"] <- (poverty$X[poverty$X.1 == "DE"] + 10000) 
poverty$X[poverty$X.1 == "DC"] <- (poverty$X[poverty$X.1 == "DC"] + 11000) 
poverty$X[poverty$X.1 == "FL"] <- (poverty$X[poverty$X.1 == "FL"] + 12000) 
poverty$X[poverty$X.1 == "GA"] <- (poverty$X[poverty$X.1 == "GA"] + 13000) 
poverty$X[poverty$X.1 == "HI"] <- (poverty$X[poverty$X.1 == "HI"] + 15000) 
poverty$X[poverty$X.1 == "ID"] <- (poverty$X[poverty$X.1 == "ID"] + 16000) 
poverty$X[poverty$X.1 == "IL"] <- (poverty$X[poverty$X.1 == "IL"] + 17000) 
poverty$X[poverty$X.1 == "IN"] <- (poverty$X[poverty$X.1 == "IN"] + 18000) 
poverty$X[poverty$X.1 == "IA"] <- (poverty$X[poverty$X.1 == "IA"] + 19000) 
poverty$X[poverty$X.1 == "KS"] <- (poverty$X[poverty$X.1 == "KS"] + 20000) 
poverty$X[poverty$X.1 == "KY"] <- (poverty$X[poverty$X.1 == "KY"] + 21000) 
poverty$X[poverty$X.1 == "LA"] <- (poverty$X[poverty$X.1 == "LA"] + 22000) 
poverty$X[poverty$X.1 == "ME"] <- (poverty$X[poverty$X.1 == "ME"] + 23000) 
poverty$X[poverty$X.1 == "MD"] <- (poverty$X[poverty$X.1 == "MD"] + 24000) 
poverty$X[poverty$X.1 == "MA"] <- (poverty$X[poverty$X.1 == "MA"] + 25000)
poverty$X[poverty$X.1 == "MI"] <- (poverty$X[poverty$X.1 == "MI"] + 26000) 
poverty$X[poverty$X.1 == "MN"] <- (poverty$X[poverty$X.1 == "MN"] + 27000) 
poverty$X[poverty$X.1 == "MS"] <- (poverty$X[poverty$X.1 == "MS"] + 28000) 
poverty$X[poverty$X.1 == "MO"] <- (poverty$X[poverty$X.1 == "MO"] + 29000) 
poverty$X[poverty$X.1 == "MT"] <- (poverty$X[poverty$X.1 == "MT"] + 30000) 
poverty$X[poverty$X.1 == "NE"] <- (poverty$X[poverty$X.1 == "NE"] + 31000) 
poverty$X[poverty$X.1 == "NV"] <- (poverty$X[poverty$X.1 == "NV"] + 32000) 
poverty$X[poverty$X.1 == "NH"] <- (poverty$X[poverty$X.1 == "NH"] + 33000) 
poverty$X[poverty$X.1 == "NJ"] <- (poverty$X[poverty$X.1 == "NJ"] + 34000) 
poverty$X[poverty$X.1 == "NY"] <- (poverty$X[poverty$X.1 == "NY"] + 36000)
poverty$X[poverty$X.1 == "NC"] <- (poverty$X[poverty$X.1 == "NC"] + 37000) 
poverty$X[poverty$X.1 == "ND"] <- (poverty$X[poverty$X.1 == "ND"] + 38000)
poverty$X[poverty$X.1 == "OH"] <- (poverty$X[poverty$X.1 == "OH"] + 39000)
poverty$X[poverty$X.1 == "OK"] <- (poverty$X[poverty$X.1 == "OK"] + 40000)
poverty$X[poverty$X.1 == "OR"] <- (poverty$X[poverty$X.1 == "OR"] + 41000) 
poverty$X[poverty$X.1 == "PA"] <- (poverty$X[poverty$X.1 == "PA"] + 42000) 
poverty$X[poverty$X.1 == "RI"] <- (poverty$X[poverty$X.1 == "RI"] + 44000) 
poverty$X[poverty$X.1 == "SC"] <- (poverty$X[poverty$X.1 == "SC"] + 45000) 
poverty$X[poverty$X.1 == "SD"] <- (poverty$X[poverty$X.1 == "SD"] + 46000) 
poverty$X[poverty$X.1 == "TN"] <- (poverty$X[poverty$X.1 == "TN"] + 47000) 
poverty$X[poverty$X.1 == "TX"] <- (poverty$X[poverty$X.1 == "TX"] + 48000) 
poverty$X[poverty$X.1 == "UT"] <- (poverty$X[poverty$X.1 == "UT"] + 49000) 
poverty$X[poverty$X.1 == "VT"] <- (poverty$X[poverty$X.1 == "VT"] + 50000) 
poverty$X[poverty$X.1 == "VA"] <- (poverty$X[poverty$X.1 == "VA"] + 51000) 
poverty$X[poverty$X.1 == "WA"] <- (poverty$X[poverty$X.1 == "WA"] + 53000) 
poverty$X[poverty$X.1 == "WV"] <- (poverty$X[poverty$X.1 == "WV"] + 54000) 
poverty$X[poverty$X.1 == "WI"] <- (poverty$X[poverty$X.1 == "WI"] + 55000) 
poverty$X[poverty$X.1 == "WY"] <- (poverty$X[poverty$X.1 == "WY"] + 56000) 
```


```{r}
poverty <- poverty |>
  mutate(countyFIPS = X)
```

```{r}
poverty <- poverty |>
  subset(countyFIPS!=0)
```
```{r}
count_deaths <- count_deaths |>
  subset(countyFIPS != 0)
```

```{r}
dataset <- poverty |>
  inner_join(count_deaths)
```

```{r}
dataset <- dataset |>
  mutate(Poverty_Count = X.3, Poverty_Perc = as.numeric(X.6)) |>
  select(countyFIPS, County_Name, final_count, final_deaths, Poverty_Count, Poverty_Perc)
```


The population data was assembled outside of this markdown file as the dataset was well over 
100 MB and needed to be further cleaned prior to its upload to github. Details of how this was done can be seen below:

```{r eval=FALSE}
data <- read.csv("cc-est2019-alldata.csv")
```
```{r eval=FALSE}
data <- data |>
  select(STATE, COUNTY, STNAME, CTYNAME, YEAR, AGEGRP, TOT_POP, WA_MALE, WA_FEMALE) |>
  filter(AGEGRP == 0, YEAR == 12) |>
  mutate(STATE = STATE *1000) |>
  mutate(COUNTY = COUNTY + STATE)
```
```{r eval=FALSE}
data2 <- data |>
  select(COUNTY, CTYNAME, TOT_POP, WA_MALE, WA_FEMALE) |>
  mutate(NWhite = TOT_POP - (WA_MALE + WA_FEMALE),
         countyFIPS = COUNTY) |>
  select(countyFIPS, CTYNAME, TOT_POP, NWhite) |>
  mutate(perc_n_white = NWhite/TOT_POP)
```
```{r eval=FALSE}
write.csv(data2, "C:\\Users\\chaseyager\\Classes\\GOV50\\final_project\\gov50-final\\data_clean\\pop.csv")
```
```{r}
pop <- read.csv("data/pop.csv")
```
```{r}
dataset <- dataset |>
  inner_join(pop, by = "countyFIPS") |>
  select(countyFIPS, County_Name, final_count, final_deaths, Poverty_Count, Poverty_Perc, TOT_POP, NWhite, perc_n_white)
```


## Data 
Sources of the data can be found at the bottom of the page. For poverty and race data all information came from the
US Census. For COVID infections and death, data was taken from USA Facts which collects government data and in this
case collected all information directly from each states' reports.

The research method used most closely resembles a cross-sectional approach, where the treatment is the rates of
poverty, and the outcome is the rate of incidence and the mortality rates of COVID-19 in the different groups. This
study would fall under this category because of the way in which the data was measured. The results of this study
are based on a snapshot of the outcomes and characteristics at a certain time, meaning there is not a before and an
after or a time dimension of our study. The date from which the data was collected is October 14th, 2022, for both
cases and deaths, as this is the latest date in the dataset and provides a running total of the cases and deaths for
the entirety of the pandemic. Additionally, information on the racial breakdown of counties taken in July of 2019
were used as a potential confounding variable to see if minority groups were additionally disproportionately
impacted by the disease. This study seeks to determine if there are preexisting differences in infection and
mortality for COVID-19 that can be attributed to preexisting differences in the populations.

The dependent variable for the study is the mortality rate of COVID-19, which was calculated by dividing the total
number of deaths on the aforementioned date due to the disease by the total number of cases at that same time. The
reason this variable was used rather than the absolute number of cases or deaths is illustrated by the following
graphs:

```{r}
pov_count_graph <- dataset |>
  ggplot(mapping = aes (x = Poverty_Perc, y = final_count)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue") +
  scale_y_log10() +
  labs(x = "Percent in Poverty by County", y = "Number of Confirmed Cases", title = "Comparison Between Poverty Rate and COVID Cases by County")
pov_count_graph
```
```{r}
pov_death_graph <- dataset |>
  ggplot(mapping = aes (x = Poverty_Perc, y = final_deaths)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue") +
  scale_y_log10() +
  labs(x = "Percent in Poverty by County", y = "Number of Confirmed Deaths", title = "Comparison Between Poverty Rate and COVID Deaths by County")

pov_death_graph
```


As can be seen, there is a slight negative correlation between poverty rates and the absolute level of cases and
deaths which would appear to go against the hypothesis. But this relationship is vulnerable to differences in
population size within counties that may alter the results. If poorer groups, often in more rural settings, have
smaller populations, then the absolute number of cases will similarly be smaller. It is for this reason that
mortality rates were used, as this provides an average for the severity of the cases, regardless of population size.
The dependent variable is summarized in the following graph:

```{r}
dataset <- dataset |>
  mutate(per_fatal = 100*(final_deaths/final_count),
         perc_n_white = 100*(perc_n_white))
```
```{r}
dataset |>
  ggplot(mapping = aes(y = per_fatal)) +
  geom_boxplot() +
  labs(title = "Distribution of Mortality Rates (deaths/cases)", y = "Deaths/Cases")
```
As can be seen. the middle 50% of the dependent variable is located at the lower end, falling between roughly 1.2%
and 1.8% for mortality rate, and the median being located at slightly over 1.25%.

## Results
The results of this study can be seen below represented in two separate graphs. The first graph is a a presentation 
of our regression results comparing our independent variable, which was the percentage of people in poverty by
county in the US, and our dependent variable, which is the mortality rates by county.

```{r}
perc_comp_fig <- dataset |>
  ggplot(mapping = aes(x = Poverty_Perc, y = per_fatal)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Percentage in Poverty Compared to Percentage of Cases that were Fatal", x = "Poverty Percentage", y = "Percentage of Cases that were Fatal")
perc_comp_fig
```
```{r}
fit3 <- lm(per_fatal ~ Poverty_Perc, data = dataset) 

fit3 |>
  broom::tidy() |>
  select(term, estimate) |>
  knitr::kable(col.names = c("Term", "Slope Estimate"))
```
```{r}
summary(fit3)
```

It is clear from the linear regression superimposed on the graph that there is a positive correlation between
poverty percentage and the percentage of cases that were fatal. The fit run on the same data demonstrates an
intercept of .719 and a slope of .054, with a p-value on bath variables of 2x10^-16 (effectively 0) indicating
highly significant results. This indicates that the estimated mortality rates when poverty is 0 would be
.719%, and that for every 1% increase in poverty percentage in a county, we would see a .054% increase in the
percentage of cases that were fatal. This would appear to confirm our hypothesis, however it is important to
consider confounding variable that may influence the results. 

For instance, it is possible that differential access to healthcare may be based more on racial divides than poverty
divides. It is for this reason that race data was also looked at as a possible descriptor of our results. 

```{r}
dataset |>
  ggplot(mapping = aes(x = perc_n_white, y = per_fatal)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Percentage Non-White Population Compared to Percentage of Fatal Cases", x = "Percentage Non-White",
       y = "Percentage of Cases that were Fatal")
```

```{r}
fit2 <- lm(per_fatal ~ perc_n_white, data = dataset) 

fit2 |>
  broom::tidy() |>
  select(term, estimate) |>
  knitr::kable(col.names = c("Term", "Estimated Slope"))
```
```{r}
summary(fit2)
```

Much like the previous results from our main relationship of interest, this graph demonstrates a positive 
relationship between the percentage of a counties population which is non-white, and the percentage of COVID-19
cases which were fatal. The fit run on this data gives an intercept of 1.37 and a slope of .006, again with both of
these results being highly significant as the p-value was effectively 0. This means that the
expected percentage of fatal cases when a population is entirely white is 1.37%, and that for every 1% increase in
the percentage of the population which is non-white, we expect to see a .006% increase in the percentage of cases
which were fatal. 

But to better understand which of the two variables, race or poverty, played a role in determining the fatality 
COVID-19 cases, a regression needed to be run which accounted for both variables. For this the following multiple
regression was run:

```{r}
fit <- lm(per_fatal ~ Poverty_Perc + perc_n_white, data = dataset) 

fit |>
  broom::tidy() |>
  select(term, estimate) |>
  knitr::kable(col.names = c("Term", "Estimated Slope"))
```

```{r}
fit |>
  tidy() |>
  knitr::kable()
```
The above table demonstrates the results of out multiple regression using both poverty percentage and the non-white 
percentage of a county as they pertain to the percentage of COVID cases which were fatal. As can be seen when
holding race constant, the estimated impact of poverty percentage was 0.062, which indicates that for a 1% increase
in poverty, we would expect to see a .062% increase in COVID-19 fatality rates. This result is highly significant,
as the p value was 0. This indicates that it is extremely unlikely that these results would occur randomly, meaning
poverty percentage does have a direct relationship with fatalities to COVID. Similarly, the p-value for the
percentage of the population that was nonwhite, while bigger, is still effectively
zero, indicating it too has an impact on the mortality rates of COVID cases, however the slope when holding poverty
percentage constant actually indicates an inverse relationship between percentage non-white and COVID fatalities.
This means that, if poverty rates are constant, a 1% increase in non-white population would lead to an expected
decrease of .005% in mortality rates. What this data goes to show, when compared to the regression run solely
between the percentage of non-white population and mortality rates, is that the likely cause of higher rates of
mortality among minority dominant populations is the result of higher poverty rates in those populations, rather
than difference in race itself. This is to say that poverty is a confounding variable in the relationship between
race and COVID mortality. This does not come as a surprise, but does speak to the socioeconomic differences between
racial groups in the US.

There are confounding variables which should be considered that may contribute to both poverty and COVID deaths,
with the most obvious being the presence of preexisting medical conditions that both prevent someone from working
and decrease the strength of their immune system; however, rates of these instances are likely too small to account
for the strength of the relationship between poverty and COVID mortality. Further, access to medical care could be
lower in rural settings, and poverty rates could be higher there contributing to both poverty and mortality, however
rural and urban centers were not weighted any differently in the study, as the study was done by county. There are
considerably more rural counties than urban counties in the study as most of the US landmass is rural, meaning urban
counties would have less impact on the average result as they would be balanced out by the larger number of rural
counties. As such, this variable likely would not have a strong enough impact on our results to compromise the
study. Given higher poverty rates likely contribute to lower rates of insurance, less access to quality medical
care, and an inability to take time away from work, which would in turn contribute to higher rates of infection and
lower rates of treatment for COVID-19, I believe these results indicate a causal relationship between COVID
mortality rates and poverty. 


## Conclusion
To recap, the results of our study, accounting for racial differences in counties, demonstrated an estimated
relationship of .062 between poverty and COVID mortality. This means that with a 1% increase in poverty, we expect
to see a .062% increase in COVID mortality rates. Further, this result was highly significant, with a p value of
zero, indicating it is highly unlikely that these results would have occurred in a random sample. 

Despite the strength of the results, there are still some limitations to the study. While the external validity of
the study was strong as it used the entire US as a sample group, the internal validity is harder to verify as there
could be a number of other variables which may have impacted our results as the treatment and control groups could
have differed in other ways that were not recorded in the data. In the future it would improve the study to control
for preexisting conditions that may contribute to both poverty and mortality, and to control for the population
density of counties which may have contributed to the rates of COVID exposure. Further, the study could be refined
to offer a clearer picture of what aspects of poverty contribute most to the higher mortality rates. Running a
regression that also included a breakdown of people classified as essential workers, people without health
insurance, and people unable to take time off from work would provide additional insights into how best to mitigate
the negative impacts of poverty on health.

## Sources
Poverty Data: [US Census](https://www.census.gov/data/datasets/2020/demo/saipe/2020-state-and-county.html )

Infection and Death Rates: [USA Facts](https://usafacts.org/visualizations/coronavirus-covid-19-spread-map/ )

Population Race Data: [US Census](https://www.census.gov/data/tables/time-series/demo/popest/2010s-counties-detail.html )
